Constraining gravitational wave amplitude birefringence with GWTC-3

### 2022-11-19 src/tex/ms.tex
Update ms.tex
New content

\author{Thomas C.K. Ng}
\email{thomas.ng@link.cuhk.edu.hk}
\affiliation{Department of Physics and Institute of Theoretical Physics, The Chinese University of Hong Kong, Shatin, Hong Kong}

@@

% Previous studies
Previous studies have constrained amplitude birefringence by performing Bayesian inference.
In \citet{Maria_2021}, they considered the distribution of observed inclination of the GW events in GWTC-2, the second LIGO-Virgo catalog,
and use the distribution to constrain the amplitude birefringence effect with the assumption of frequency independence in GW amplitude birefringence.
In \citet{Yamada_2020} and \citet{Wang_2021}, they both modeled the waveform under birefringence effect with a simplify assumption and
performed Bayesian inference on the events in GWTC-1, the first LIGO-Virgo catalog.

% What's new?
In this paper, we used a birefringence model with higher order terms which include the frequency dependence of amplitude birefringence
to perform a more accurate parameter estimation (PE).
We also performed Bayesian inference on events in GWTC-3, the third LIGO-Virgo catalog.
We included all binary black hole merger events with smallest false alarm rate $\leq1\mathrm{yr^{-1}}$, which are listed in \citet{GWTC_3_population}.

@@

% Parameter estimation
To constrain $\kappa$, we perform PE with data from the third LIGO-Virgo catalog \citep{GWTC-2.1, GWTC-3}, GWTC-3, using Bilby,
a Bayesian toolkit for GW data analysis which is able to calculate posteriors of GW parameters based on interferometer data
and priors for the parameters in question \citep{Bilby}. 
We set the prior on $\kappa$ to a uniform distribution between $-1$ and $1$
\footnote{For \textbf{[some events]}, a uniform distribution between $-0.5$ and $0.5$ was used instead. This does not affect the posterior distribution.},
and perform PE using the modified waveforms.
In both case, we use the same PE configuration as in \citet{GWTC-2.1, GWTC-3} for each individual events, excluding the calibration of the detectors.
We also perform test to check if the PE could recover the PE done by LIGO in \citet{GWTC-2.1, GWTC-3}.
We set the prior on $\kappa$ to be a $\delta$ function at $0$, which is the same as using GR.
These control tests result can be found in the data release.

@@

\subsection{Result on GWTC-3}
% Combined plot

@@

@@

@@

\begin{figure}[h]
    \script{GW150914_corner.py}
    \includegraphics[width=\columnwidth]{figures/GW150914_corner.pdf}
    \caption{
        The posterior of $\kappa$, luminosity distance $d_L$ and $\cos{\iota}$ for GW150914.
    The three colors in the plot are the PE with GR done by LIGO without cosmological reweighing \citep{GWTC-2.1, GWTC-3},
    the PE done by us with the frequency independent birefringence and the frequency dependent birefringence respectively.
    Note that there is no posterior of $\kappa$ for the PE from LIGO, as the LIGO PE is based on GR,
    which does not suggest GW amplitude birefringence. The plot shows that frequency independent birefringence create a degeneracy between $\kappa$ and $\iota$,
    while the frequency dependent birefringence can break it.
    }
    \label{fig:GW150914_corner}
\end{figure}

Consider GW150914, the first GW detected by LIGO. In figure \ref{fig:GW150914_corner}, we show the posteriors of the parameters of GW150914.
With the frequency independent birefringence model, the posteriors for $\cos\iota$ look different from the posteriors assuming GR.
This is because, for a nonprecessing system, there is a degeneracy between $\kappa$ and $\iota$ if the frequency dependence is not included.
To reconstruct the amplitude ratio from the interferometer data, an $\iota$ representing a more face-off inspiral can pair with a positive $\kappa$,
or an $\iota$ representing a more face-on inspiral with a negative $\kappa$.
As a result,many pairs of $\iota$ and $\kappa$ can be plausible with the frequency independence.

On the other hand, with the frequency-dependent birefringence model, the posterior looks similar to the GR posterior.
This is because the degeneracy was broken by the frequency dependence, as the effect of the birefringence will be different from the one of changing $\iota$.
In this case, the posteriors can recover the GR distributions for the binary parameters,
and the most probable value of $\kappa$ is close to $0$, which means the birefringence is weak or absent, and GR can be recovered.

@@

% BNS

% Higher order terms

% More observations with higher SNR

>>>>

\author{Thomas C. K. Ng}
\email{thomas.ng@link.cuhk.edu.hk}
\affiliation{Department of Physics, The Chinese University of Hong Kong, Shatin, Hong Kong}

@@

% Previous studies
Previous studies have constrained amplitude birefringence by performing Bayesian inference.
In \citet{Maria_2021}, they considered the distribution of observed inclination of the GW events in GWTC-2, the second GW transient catalog, and use the distribution to constrain the amplitude birefringence effect with the assumption of frequency independence in GW amplitude birefringence.
In \citet{Yamada_2020} and \citet{Wang_2021}, they both modeled the waveform under birefringence effect with a simplify assumption and performed Bayesian inference on the events in GWTC-1, the first GW transient catalog.

% What's new?
In this paper, we used a birefringence model with higher order terms which include the frequency dependence of amplitude birefringence to perform a more accurate parameter estimation (PE).
We also performed Bayesian inference on events in GWTC-3, the third GW transient catalog.
We included 70 binary black hole merger events with smallest false alarm rate $\leq1\mathrm{yr^{-1}}$, which are listed in \citet{GWTC_3_population}.\footnote{GW190720 is included in list, however, we could not successfully perform PE with it. Details on GW190720 will be mention in Discussion.}

@@

% Parameter estimation
The major objective of this study is to constrain GW amplitude birefringence, which is represented by the opacity parameter $\kappa$.
To constrain $\kappa$, we perform PE with data from the third GW transient catalog \citep{GWTC-2.1, GWTC-3}, GWTC-3, using Bilby, a Bayesian toolkit for GW data analysis which is able to calculate posteriors of GW parameters based on interferometer data and priors for the parameters in question \citep{Bilby}. 
We set the prior on $\kappa$ to a uniform distribution between $-1$ and $1$, and perform PE using the modified waveforms.
In both case, we use the same PE configuration as in \citet{GWTC-2.1, GWTC-3} for each individual events.
We also perform test to check if the PE could recover the PE done by LIGO-Virgo-KAGRA Collaboration (LVK) in \citet{GWTC-2.1, GWTC-3}.
We set the prior on $\kappa$ to be a $\delta$ function at $0$, which is the same as using GR.
These control tests result can be found in the data release.

@@

\subsection{Result on GWTC-3}
% Stacked kappa posterior plot
In figure \ref{fig:kappa_stacked}, we show the individual posterior of $\kappa$ of each events.

\begin{figure}[h]
    \script{kappa_stacked.py}
    % \includegraphics[width=\columnwidth]{figures/kappa_stacked.pdf}
    \caption{
        The posterior of $kappa$ for all 70 events included in this study.
        Each color represent a different event.
        These posterior are used to calculate the constraint on $\kappa$ in the following.
    }
    \label{fig:kappa_stacked}
\end{figure}

@@

We assume $\kappa$ is different for each event, and the population distribution of $\kappa$ is a Gaussian distribution with mean $\mu$ and standard deviation $\sigma$.
We then performed hierarchical modeling to construct the likelihood function of $\mu$ and $\sigma$ given interferometer data.
To sample the posterior distribution of $\mu$ and $\sigma$, we used a generic sampling package, flowMC \citep{flowMC}, with prior of $\mu$ and $\sigma$ being uniform distributions.
The posterior of $\mu$ and $\sigma$ are shown in figure \ref{fig:corner_Gaussian}.

% Corner plot of mean and standard deviation of kappa

\begin{figure}[h]
    \script{corner_Gaussian.py}
    % \includegraphics[width=\columnwidth]{figures/corner_Gaussian.pdf}
    \caption{
        The posterior of $\mu$ and $\sigma$ of the population distribution of $\kappa$.
    }
    \label{fig:corner_Gaussian}
\end{figure}

@@

\begin{figure}[h]
    \script{violin_kappa.py}
    % \includegraphics[width=\columnwidth]{figures/violin_kappa.pdf}
    \caption{
        The violin plot shows the posterior of $kappa$ for all 70 events included in this study.
        Each violin represents a different event.
        The violins are sorted by the quotient of median and standard deviation of the posterior.
    }
    \label{fig:violin_kappa}
\end{figure}

@@

\begin{figure}[h]
    \script{corner_GW150914.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW150914.pdf}
    \caption{
        The posterior of $\kappa$, luminosity distance $d_L$ and $\cos{\iota}$ for GW150914.
        The three colors in the plot are the PE with GR done by LVK without cosmological reweighing \citep{GWTC-2.1, GWTC-3}, the PE done by us with the frequency independent birefringence and the frequency dependent birefringence respectively.
        Note that there is no posterior of $\kappa$ for the PE from LVK, as the LVK PE is based on GR, which does not suggest GW amplitude birefringence.
        The plot shows that frequency independent birefringence create a degeneracy between $\kappa$ and $\iota$, while the frequency dependent birefringence can break it.
    }
    \label{fig:corner_GW150914}
\end{figure}

Consider GW150914, the first GW detected by LIGO.
In figure \ref{fig:GW150914_corner}, we show the posteriors of the parameters of GW150914.
With the frequency independent birefringence model, the posteriors for $\cos\iota$ look different from the posteriors assuming GR.
This is because, for a nonprecessing system, there is a degeneracy between $\kappa$ and $\iota$ if the frequency dependence is not included.
To reconstruct the amplitude ratio from the interferometer data, an $\iota$ representing a more face-off inspiral can pair with a positive $\kappa$, or an $\iota$ representing a more face-on inspiral with a negative $\kappa$.
As a result, many pairs of $\iota$ and $\kappa$ can be plausible with the frequency independence.

On the other hand, with the frequency-dependent birefringence model, the posterior looks similar to the GR posterior.
This is because the degeneracy was broken by the frequency dependence, as the effect of birefringence will be different from the one of changing $\iota$.
In this case, the posteriors can recover the GR distributions for the binary parameters,
and the most probable value of $\kappa$ is close to $0$, which means the birefringence is weak or absent, and GR can be recovered.

% Case 2: GW190521 (massive BBH)

\begin{figure}[h]
    \script{corner_GW190521.py}
    % \includegraphics[width=\columnwidth]{figures/corner_GW190521.pdf}
    \caption{
        The posterior of $\kappa$, luminosity distance $d_L$ and $\cos{\iota}$ for GW190521.
        The colors in the plot are the PE with GR done by LVK without cosmological reweighing \citep{GWTC-2.1, GWTC-3} and the PE done by us with the frequency dependent birefringence.
        The plot shows that frequency dependent birefringence cannot break the degeneracy between $\kappa$ and $\iota$.
    }
    \label{fig:corner_GW190521}
\end{figure}

Consider GW190521, the most massive binary black hole merger in the events we included.
The degeneracy between $\kappa$ and $\iota$ cannot be broken by the frequency dependent birefringence model, as shown in figure \ref{fig:corner_GW190521}.
This is because the frequency range of the signal is much smaller due to the large mass of the binary.
The effect of birefringence at different frequency within the range is similar, thus the degeneracy cannot be broken.

@@

% Special event 1: GW190720

% Special event 2: GW200129

% Future work

% BNS
A future work is to perform PE on binary neutron star mergers, such as GW170817.
The frequency range of the signal is much larger compared to the binary black hole mergers, thus the difference of the effect of birefringence at different frequency within the range can be more significant.
This can allow us to further constrain the birefringence effect, and the beyond-GR theories that predict it.
However, performing PE on binary neutron star mergers requires much more computational resources.
Therefore, we may need to wait for future PE methods and tools to further our work.

% More observations with higher SNR
Another future work is to apply the same method to more events with higher signal-to-noise ratio (SNR).
LVK will release more events with higher SNR in the future.
